---
title: "curc_work"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,      # Hide code
  warning = FALSE,   # Suppress warnings
  message = FALSE,   # Suppress messages
  error = FALSE      # Suppress errors (optional, use TRUE to debug)
)
```

# General first data pull 
```{r, include=FALSE}
source("setup.R")

# Download the meta-data for all sites
meta_data <- snotel_info()

# View the first few rows to understand the structure
head(meta_data)

id = 438 # Deadman Hill = 438
# Download data for a specific site and return as a data frame
df <- snotel_download(site_id = id, internal = TRUE)

# View the first few rows of the downloaded data
head(df)

# make this column date objects
df$date <- as.Date(df$date)

# Make columns for day of water year and water year
df <- df %>%
  mutate(
    water_year_start = if_else(month(date) >= 10,
                               make_date(year(date), 10, 1),
                               make_date(year(date) - 1, 10, 1)),
    day_of_water_year = as.integer(date - water_year_start) + 1,
    water_year = if_else(month(date) >= 10,
                         year(date)+1,
                         year(date),
  )) %>%
  select(-water_year_start)

df %>% select(date, water_year, day_of_water_year) %>% head()

# view new columns
print(df$day_of_water_year)
```

# Computations for depth x SWE
```{r}
# getting a better estimate than Mizukami (will just be comparison for what we do)
# density(ps) = denswater(pw) (swe/depth)
# take an average value of the densities over around 15 years (up until 2020)
  # apply it to the last 5 years individually to evaluate how good that average is 
# compare to the 164.7 kg/m^3
df <- df %>%
  mutate(
    snow_density = case_when(
      is.na(snow_depth) | is.na(snow_water_equivalent) ~ NA_real_,
      snow_depth <= 0 ~ NA_real_,
      TRUE ~ (snow_water_equivalent / snow_depth) * 1000  # kg/m^3
    )
  )

library(openxlsx)
write.xlsx(df, "SNOTEL_deadman_hill.xlsx")
```

# 15-year graph with trend
```{r, echo=FALSE}
# one fitted avg density line from Dec 1 - Mar 1, Mar 1 - Apr 30 ON TOP of the 15 years 

# ----------------------------
# 1) Keep full water year data (2005–2020)
# ----------------------------
df_0520 <- df %>%
  filter(water_year >= 2005, water_year <= 2020) %>%
  filter(!is.na(snow_density)) %>%
  arrange(water_year, day_of_water_year)

# ----------------------------
# 2) Average density by day-of-water-year across years
# ----------------------------
avg_by_day <- df_0520 %>%
  group_by(day_of_water_year) %>%
  summarise(avg_density = mean(snow_density, na.rm = TRUE), .groups = "drop") %>%
  arrange(day_of_water_year)

# ----------------------------
# 3) Define key day-of-water-year cut points (non-leap reference)
# WY starts Oct 1 = day 1
# Dec 1 = 62, Mar 1 = 152, Apr 30 = 212
# ----------------------------
dec1 <- 62
mar1 <- 152
apr30 <- 212

# Fit datasets (only where you want the trend modeled)
fit1_data <- avg_by_day %>% filter(day_of_water_year >= dec1, day_of_water_year <= mar1)
fit2_data <- avg_by_day %>% filter(day_of_water_year >= mar1, day_of_water_year <= apr30)

# Models
fit1 <- lm(avg_density ~ day_of_water_year, data = fit1_data)
fit2 <- lm(avg_density ~ day_of_water_year, data = fit2_data)

# ----------------------------
# 4) One continuous piecewise prediction (NO BREAKS)
# ----------------------------
pred_piecewise <- tibble(
  day_of_water_year = seq(dec1, apr30)
) %>%
  mutate(
    fit_density = if_else(
      day_of_water_year <= mar1,
      as.numeric(predict(fit1, newdata = .)),
      as.numeric(predict(fit2, newdata = .))
    )
  ) %>%
  filter(is.finite(fit_density))

# ----------------------------
# 5) Month breaks/labels ONLY for Dec–Apr window
# ----------------------------
month_breaks  <- c(62, 93, 124, 152, 183, 212)   # Dec 1, Jan 1, Feb 1, Mar 1, Apr 1, Apr 30
month_labels  <- c("Dec","Jan","Feb","Mar","Apr","Apr 30")

# ----------------------------
# 6) Plot (Dec 1–Apr 30 only)
# ----------------------------
df_0520_window <- df_0520 %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30)

ggplot(df_0520_window,
       aes(x = day_of_water_year, y = snow_density,
           group = water_year, color = factor(water_year))) +

  geom_line(alpha = 0.8) +

  # Continuous fitted line (piecewise)
  geom_line(data = pred_piecewise,
            aes(x = day_of_water_year, y = fit_density, group = 1),
            inherit.aes = FALSE,
            linewidth = 1,
            color = "black") +

  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +

  labs(
    title = "Bulk Snowpack Density by Water Year (2005–2020) — Dec 1 to Apr 30",
    x = "Month (Water Year)",
    y = "Bulk Snow Density (kg/m³)",
    color = "Water Year"
  ) +
  theme_minimal()
```

### trends over the years (for poster) with the M-P fitted line
```{r}
# empirical density model from Dec 1 - Mar 1 and Mar 1 - Apr 30 ON TOP of the years 

library(dplyr)
library(ggplot2)
library(tibble)

# ----------------------------
# 1) Keep full water year data (2005–2025)
# ----------------------------
df_0525 <- df %>%
  filter(water_year >= 2005, water_year <= 2025) %>%
  filter(!is.na(snow_density)) %>%
  arrange(water_year, day_of_water_year)

# ----------------------------
# 2) Define key day-of-water-year cut points (non-leap reference)
# WY starts Oct 1 = day 1
# Dec 1 = 62, Mar 1 = 152, Apr 30 = 212
# ----------------------------
dec1 <- 62
mar1 <- 152
apr30 <- 212

# ----------------------------
# 3) Build piecewise empirical model
# Dec 1–Mar 1:   density = 164.7 + 0.4 * t_WY
# Mar 1–Apr 30:  density = -90.7 + 2.1 * t_WY
# ----------------------------
model_piecewise <- tibble(
  day_of_water_year = seq(dec1, apr30)
) %>%
  mutate(
    model_density = case_when(
      day_of_water_year <= mar1 ~ 164.7 + 0.4 * day_of_water_year,
      day_of_water_year >  mar1 ~ -90.7 + 2.1 * day_of_water_year
    )
  )

# ----------------------------
# 4) Month breaks/labels ONLY for Dec–Apr window
# ----------------------------
month_breaks <- c(62, 93, 124, 152, 183, 212)
month_labels <- c("Dec","Jan","Feb","Mar","Apr","Apr 30")

# ----------------------------
# 5) Plot window and highlight years
# ----------------------------
df_0525_window <- df_0525 %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30) %>%
  mutate(
    year_group = case_when(
      water_year == 2011 ~ "2011",
      water_year == 2012 ~ "2012",
      TRUE ~ "Other"
    )
  )

ggplot(df_0525_window,
       aes(x = day_of_water_year, y = snow_density,
           group = water_year, color = year_group)) +

  geom_line(alpha = 0.8) +

  # Piecewise empirical model
  geom_line(data = model_piecewise,
            aes(x = day_of_water_year, y = model_density, group = 1),
            inherit.aes = FALSE,
            linewidth = 1.2,
            color = "red") +

  scale_color_manual(
    values = c(
      "Other" = "grey75",
      "2011" = "#2259E0",
      "2012" = "#38E022"
    ),
    breaks = c("2011", "2012", "Other"),
    labels = c("2011", "2012", "Other years:\n 2005-2010,\n 2013-2025")
  ) +

  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +

  labs(
    title = "Bulk Snowpack Density by Water Year (2005–2025) at Deadman Hill SNOTEL",
    subtitle = "Two-part linear snowpack density equations (red line) derived from Mizukami and Perica (2008)",
    x = "Month (Water Year)",
    y = "Bulk Snow Density (kg/m³)",
    color = "Water Year"
  ) +
  theme_minimal()
```
**Figure 1.** Seasonal evolution of bulk snowpack density at the Deadman Hill SNOTEL site for water years 2005–2025, restricted to December 1 through April 30. Water years 2011 and 2012 are highlighted for comparison, while all remaining years are shown in gray. The red line shows a piecewise empirical density model using 𝜌=164.7+0.4" " 𝑡_𝑊𝑌from December 1 to March 1 and 𝜌=−90.7+2.1" " 𝑡_𝑊𝑌from March 1 to April 30 (1). 

# Daily SWE, depth, and density 
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)

# ----------------------------
# Daily change variables
# ----------------------------
df_daily <- df %>%
  arrange(water_year, date) %>%
  group_by(water_year) %>%
  mutate(
    d_swe   = snow_water_equivalent - lag(snow_water_equivalent),
    d_depth = snow_depth - lag(snow_depth),
    d_dens  = snow_density - lag(snow_density)
  ) %>%
  ungroup()

# ----------------------------
# Keep WY 2011 and 2012, Dec 1–Apr 30
# ----------------------------
plot_1112 <- df_daily %>%
  filter(water_year %in% c(2011, 2012),
         day_of_water_year >= 62,
         day_of_water_year <= 212)

# ----------------------------
# SWE and Depth plot
# ----------------------------
ggplot(plot_1112, aes(x = date)) +
  geom_line(aes(y = snow_water_equivalent, color = "SWE"), linewidth = 0.9) +
  geom_line(aes(y = snow_depth, color = "Depth"), linewidth = 0.9, linetype = "dashed") +
  facet_wrap(~water_year, ncol = 1, scales = "free_x") +
  labs(
    title = "Daily SWE and Snow Depth, WY 2011 and 2012",
    subtitle = "Comparing daily increases and decreases in SWE and depth",
    x = "Date",
    y = "Value (mm)",
    color = ""
  ) +
  theme_minimal()


### Density plot for the same years
ggplot(plot_1112, aes(x = date, y = snow_density)) +
  geom_line(linewidth = 0.9, color = "black") +
  geom_hline(yintercept = 164.7, linetype = "dotted", color = "red") +
  facet_wrap(~water_year, ncol = 1, scales = "free_x") +
  labs(
    title = "Daily Bulk Snow Density, WY 2011 and 2012",
    subtitle = "Red dotted line = average fresh snow density (164.7 kg/m³)",
    x = "Date",
    y = "Bulk Snow Density (kg/m³)"
  ) +
  theme_minimal()
```

# Daily changes

---

## 2) Plot daily changes directly to see the pattern better

**density decreases while depth increases**, then later **depth settles back down**.

```{r}
changes_1112 <- df_daily %>%
  filter(water_year %in% c(2011, 2012),
         day_of_water_year >= 62,
         day_of_water_year <= 212)

ggplot(changes_1112, aes(x = date)) +
  geom_hline(yintercept = 0, color = "grey60") +
  geom_col(aes(y = d_depth), fill = "skyblue3", alpha = 0.7) +
  geom_line(aes(y = d_dens / 10), color = "darkred", linewidth = 0.9) +
  facet_wrap(~water_year, ncol = 1, scales = "free_x") +
  labs(
    title = "Daily Change in Snow Depth with Density Change Overlay",
    subtitle = "Bars = change in depth (mm); red line = change in density scaled by 10",
    x = "Date",
    y = "Daily change"
  ) +
  theme_minimal()
```

### How to read that plot
- **Large positive depth bars** with a **drop in density** suggest **new, low-density snowfall**
- After that, **depth may decline gradually** while **density rises**, which suggests **settling/compaction**
- The reverse pattern is usually less consistent with fresh snowfall

---

## 3) Classify likely snowfall vs settling events

compare each day’s density to a **fresh snow reference density**.
```{r}
fresh_snow_density <- 164.7

df_events <- df_daily %>%
  mutate(
    density_dev_fresh = snow_density - fresh_snow_density,
    event_type = case_when(
      is.na(d_swe) | is.na(d_depth) | is.na(snow_density) ~ NA_character_,
      
      # likely new snowfall:
      # SWE increases, depth increases, density is near or below fresh snow density
      d_swe > 0 & d_depth > 0 & snow_density <= fresh_snow_density ~ "Likely new snowfall",
      
      # likely settling/compaction:
      # SWE stable or small change, depth decreases, density increases
      d_depth < 0 & d_dens > 0 ~ "Likely settling",
      
      # SWE increases but depth does not increase much = denser addition / rain-on-snow / mixed process
      d_swe > 0 & d_depth <= 0 ~ "SWE gain without depth gain",
      
      TRUE ~ "Other / mixed"
    )
  )
```


---

## 5) Scatterplot to distinguish snowfall vs settling

- x = **change in depth**
- y = **change in density**

That gives:
- **upper left / lower right** type structure depending on process
- **fresh snowfall** should often be **depth up, density down**
- **settling** should often be **depth down, density up**
```{r}
scatter_1112 <- df_events %>%
  filter(water_year %in% c(2011, 2012),
         day_of_water_year >= 62,
         day_of_water_year <= 212,
         !is.na(d_depth),
         !is.na(d_dens))

ggplot(scatter_1112, aes(x = d_depth, y = d_dens, color = event_type)) +
  geom_hline(yintercept = 0, color = "grey60") +
  geom_vline(xintercept = 0, color = "grey60") +
  geom_point(alpha = 0.8) +
  facet_wrap(~water_year) +
  labs(
    title = "Daily Change in Depth vs Daily Change in Density",
    subtitle = "Fresh snowfall often appears as depth increase + density decrease; settling as depth decrease + density increase",
    x = "Daily change in snow depth (mm)",
    y = "Daily change in density (kg/m³)",
    color = "Interpretation"
  ) +
  theme_minimal()
```

---

## 6) Add precipitation to confirm snowfall days

- **Precip > 0**
- **SWE increases**
- **Depth increases or stays similar**
- **Density decreases or stays near fresh snow density**

That is strong evidence for **new snowfall**
```{r}
df_events <- df_events %>%
   mutate(
     event_type_precip = case_when(
      precipitation > 0 & d_swe > 0 & d_depth >= 0 & d_dens <= 0 ~ "Likely snowfall",
      precipitation == 0 & d_depth < 0 & d_dens > 0 ~ "Likely settling",
       TRUE ~ "Other / mixed"
    )
   )
```

---

## 7) A cleaner event-based filter

This isolates the pattern:

> when density spikes down, depth spikes up, then comes down later at a different settling rate

```{r}
hypothesis_days <- df_events %>%
  filter(water_year %in% c(2011, 2012),
         day_of_water_year >= 62,
         day_of_water_year <= 212) %>%
  mutate(
    spike_pattern = case_when(
      d_dens < 0 & d_depth > 0 ~ "Density down, depth up",
      d_dens > 0 & d_depth < 0 ~ "Settling signature",
      TRUE ~ "Other"
    )
  )

ggplot(hypothesis_days, aes(x = date, y = d_depth, fill = spike_pattern)) +
  geom_col() +
  facet_wrap(~water_year, ncol = 1, scales = "free_x") +
  labs(
    title = "Days Matching the Expected Snowfall/Settling Pattern",
    subtitle = "Looking for density decreases paired with depth increases, followed by settling-type behavior",
    x = "Date",
    y = "Daily change in depth (mm)",
    fill = ""
  ) +
  theme_minimal()
```

---

## Step-by-step: modeled density → modeled depth → daily changes → difference plot
```{r}
library(dplyr)
library(ggplot2)

# ----------------------------
# 1) Filter WY 2011 and 2012
# ----------------------------
df_1112 <- df %>%
  filter(water_year %in% c(2011, 2012),
         day_of_water_year >= 62,
         day_of_water_year <= 212) %>%
  arrange(water_year, date)

# ----------------------------
# 2) Mizukami & Perica modeled density
# ----------------------------
df_1112 <- df_1112 %>%
  mutate(
    density_model = case_when(
      day_of_water_year <= 152 ~ 164.7 + 0.4 * day_of_water_year,
      day_of_water_year > 152  ~ -90.7 + 2.1 * day_of_water_year
    )
  )

# ----------------------------
# 3) Modeled snow depth
# SWE (mm) / density (kg/m³) → depth (mm)
# depth = SWE / density * 1000
# ----------------------------
df_1112 <- df_1112 %>%
  mutate(
    depth_model = case_when(
      is.na(snow_water_equivalent) | is.na(density_model) ~ NA_real_,
      density_model <= 0 ~ NA_real_,
      TRUE ~ (snow_water_equivalent / density_model) * 1000
    )
  )

# ----------------------------
# 4) Daily changes (observed + modeled)
# ----------------------------
df_1112 <- df_1112 %>%
  group_by(water_year) %>%
  mutate(
    d_depth_obs   = snow_depth - lag(snow_depth),
    d_depth_model = depth_model - lag(depth_model)
  ) %>%
  ungroup()

# ----------------------------
# 5) Difference in daily change
# ----------------------------
df_1112 <- df_1112 %>%
  mutate(
    d_depth_diff = d_depth_obs - d_depth_model
  )

# ----------------------------
# 6) Plot: Daily Change in Snow Depth
# ----------------------------
ggplot(df_1112, aes(x = date)) +
  geom_hline(yintercept = 0, color = "black") +

  geom_line(aes(y = d_depth_obs, color = "Observed"), linewidth = 0.9) +
  geom_line(aes(y = d_depth_model, color = "Modeled"), linewidth = 0.9) +

  facet_wrap(~water_year, ncol = 1, scales = "free_x") +

  labs(
    title = "Daily Change in Snow Depth (Observed vs Modeled)",
    subtitle = "Modeled depth derived from Mizukami & Perica density model",
    x = "Date",
    y = "Daily change in snow depth (mm)",
    color = ""
  ) +
  theme_minimal()

ggplot(df_1112, aes(x = date, y = d_depth_diff)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_line(color = "purple", linewidth = 1) +

  facet_wrap(~water_year, ncol = 1, scales = "free_x") +

  labs(
    title = "Difference in Daily Snow Depth Change",
    subtitle = "Observed - Modeled",
    x = "Date",
    y = "Δ Depth Difference (mm)"
  ) +
  theme_minimal()
```