---
title: "data_work"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,      # Hide code
  warning = FALSE,   # Suppress warnings
  message = FALSE,   # Suppress messages
  error = FALSE      # Suppress errors (optional, use TRUE to debug)
)
```

# General first data pull 
```{r, include=FALSE}
source("setup.R")

# Download the meta-data for all sites
meta_data <- snotel_info()

# View the first few rows to understand the structure
head(meta_data)

id = 438 # Deadman Hill = 438
# Download data for a specific site and return as a data frame
df <- snotel_download(site_id = id, internal = TRUE)

# View the first few rows of the downloaded data
head(df)

# make this column date objects
df$date <- as.Date(df$date)

# Make columns for day of water year and water year
df <- df %>%
  mutate(
    water_year_start = if_else(month(date) >= 10,
                               make_date(year(date), 10, 1),
                               make_date(year(date) - 1, 10, 1)),
    day_of_water_year = as.integer(date - water_year_start) + 1,
    water_year = if_else(month(date) >= 10,
                         year(date)+1,
                         year(date),
  )) %>%
  select(-water_year_start)

df %>% select(date, water_year, day_of_water_year) %>% head()

# view new columns
print(df$day_of_water_year)

# View a water year
library(ggplot2)
df_subset <- df %>% filter(water_year == 2025) # filter df to one year
ggplot(df_subset, aes(x = date)) +
  geom_line(aes(y = snow_water_equivalent, color = "SWE")) +
  geom_line(aes(y = precipitation_cumulative, color = "Precipitation")) +
  labs(
    title = "Snow Water Equivalent and Precipitation – Water Year 2025",
    x = "Date",
    y = "mm of water",
    color = ""
  ) +
  theme_minimal()

# Function to look at different time periods
compare_mean_swe_by_wyday <- function(
    df,
    swe_col = snow_water_equivalent,
    period1 = c(1980, 2000),
    period2 = c(2001, 2025)
) {
  
  df_proc <- df %>%
    mutate(
      period = case_when(
        water_year >= period1[1] & water_year <= period1[2] ~
          paste0(period1[1], "–", period1[2]),
        water_year >= period2[1] & water_year <= period2[2] ~
          paste0(period2[1], "–", period2[2]),
        TRUE ~ NA_character_
      )
    ) %>%
    filter(!is.na(period))
  
  swe_mean <- df_proc %>%
    group_by(period, day_of_water_year) %>%
    summarise(
      mean_swe = mean({{ swe_col }}, na.rm = TRUE),
      .groups = "drop"
    )
  
  ggplot(swe_mean, aes(x = day_of_water_year, y = mean_swe, color = period)) +
    geom_line(linewidth = 1) +
    labs(
      title = "Mean Snow Water Equivalent by Day of Water Year",
      x = "Day of Water Year",
      y = "Mean Snow Water Equivalent",
      color = "Period"
    ) +
    theme_minimal()
}

compare_mean_swe_by_wyday(df)
compare_mean_swe_by_wyday(df, swe_col=precipitation_cumulative)
```

# Computations for depth x SWE
```{r}
# getting a better estimate than Mizukami (will just be comparison for what we do)
# density(ps) = denswater(pw) (swe/depth)
# take an average value of the densities over around 15 years (up until 2020)
  # apply it to the last 5 years individually to evaluate how good that average is 
# compare to the 164.7 kg/m^3
df <- df %>%
  mutate(
    snow_density = case_when(
      is.na(snow_depth) | is.na(snow_water_equivalent) ~ NA_real_,
      snow_depth <= 0 ~ NA_real_,
      TRUE ~ (snow_water_equivalent / snow_depth) * 1000  # kg/m^3
    )
  )
```

# 15-year graph with trend
```{r, echo=FALSE}
# one fitted avg density line from Dec 1 - Mar 1, Mar 1 - Apr 30 ON TOP of the 15 years 

# ----------------------------
# 1) Keep full water year data (2005–2020)
# ----------------------------
df_0520 <- df %>%
  filter(water_year >= 2005, water_year <= 2020) %>%
  filter(!is.na(snow_density)) %>%
  arrange(water_year, day_of_water_year)

# ----------------------------
# 2) Average density by day-of-water-year across years
# ----------------------------
avg_by_day <- df_0520 %>%
  group_by(day_of_water_year) %>%
  summarise(avg_density = mean(snow_density, na.rm = TRUE), .groups = "drop") %>%
  arrange(day_of_water_year)

# ----------------------------
# 3) Define key day-of-water-year cut points (non-leap reference)
# WY starts Oct 1 = day 1
# Dec 1 = 62, Mar 1 = 152, Apr 30 = 212
# ----------------------------
dec1 <- 62
mar1 <- 152
apr30 <- 212

# Fit datasets (only where you want the trend modeled)
fit1_data <- avg_by_day %>% filter(day_of_water_year >= dec1, day_of_water_year <= mar1)
fit2_data <- avg_by_day %>% filter(day_of_water_year >= mar1, day_of_water_year <= apr30)

# Models
fit1 <- lm(avg_density ~ day_of_water_year, data = fit1_data)
fit2 <- lm(avg_density ~ day_of_water_year, data = fit2_data)

# ----------------------------
# 4) One continuous piecewise prediction (NO BREAKS)
# ----------------------------
pred_piecewise <- tibble(
  day_of_water_year = seq(dec1, apr30)
) %>%
  mutate(
    fit_density = if_else(
      day_of_water_year <= mar1,
      as.numeric(predict(fit1, newdata = .)),
      as.numeric(predict(fit2, newdata = .))
    )
  ) %>%
  filter(is.finite(fit_density))

# ----------------------------
# 5) Month breaks/labels ONLY for Dec–Apr window
# ----------------------------
month_breaks  <- c(62, 93, 124, 152, 183, 212)   # Dec 1, Jan 1, Feb 1, Mar 1, Apr 1, Apr 30
month_labels  <- c("Dec","Jan","Feb","Mar","Apr","Apr 30")

# ----------------------------
# 6) Plot (Dec 1–Apr 30 only)
# ----------------------------
df_0520_window <- df_0520 %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30)

ggplot(df_0520_window,
       aes(x = day_of_water_year, y = snow_density,
           group = water_year, color = factor(water_year))) +

  geom_line(alpha = 0.8) +

  # Continuous fitted line (piecewise)
  geom_line(data = pred_piecewise,
            aes(x = day_of_water_year, y = fit_density, group = 1),
            inherit.aes = FALSE,
            linewidth = 1,
            color = "black") +

  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +

  labs(
    title = "Bulk Snowpack Density by Water Year (2005–2020) — Dec 1 to Apr 30",
    x = "Month (Water Year)",
    y = "Bulk Snow Density (kg/m³)",
    color = "Water Year"
  ) +
  theme_minimal()
```

### trends over the years (for poster) with the M-P fitted line
```{r}
# empirical density model from Dec 1 - Mar 1 and Mar 1 - Apr 30 ON TOP of the years 

library(dplyr)
library(ggplot2)
library(tibble)

# ----------------------------
# 1) Keep full water year data (2005–2025)
# ----------------------------
df_0525 <- df %>%
  filter(water_year >= 2005, water_year <= 2025) %>%
  filter(!is.na(snow_density)) %>%
  arrange(water_year, day_of_water_year)

# ----------------------------
# 2) Define key day-of-water-year cut points (non-leap reference)
# WY starts Oct 1 = day 1
# Dec 1 = 62, Mar 1 = 152, Apr 30 = 212
# ----------------------------
dec1 <- 62
mar1 <- 152
apr30 <- 212

# ----------------------------
# 3) Build piecewise empirical model
# Dec 1–Mar 1:   density = 164.7 + 0.4 * t_WY
# Mar 1–Apr 30:  density = -90.7 + 2.1 * t_WY
# ----------------------------
model_piecewise <- tibble(
  day_of_water_year = seq(dec1, apr30)
) %>%
  mutate(
    model_density = case_when(
      day_of_water_year <= mar1 ~ 164.7 + 0.4 * day_of_water_year,
      day_of_water_year >  mar1 ~ -90.7 + 2.1 * day_of_water_year
    )
  )

# ----------------------------
# 4) Month breaks/labels ONLY for Dec–Apr window
# ----------------------------
month_breaks <- c(62, 93, 124, 152, 183, 212)
month_labels <- c("Dec","Jan","Feb","Mar","Apr","Apr 30")

# ----------------------------
# 5) Plot window and highlight years
# ----------------------------
df_0525_window <- df_0525 %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30) %>%
  mutate(
    year_group = case_when(
      water_year == 2011 ~ "2011",
      water_year == 2012 ~ "2012",
      TRUE ~ "Other"
    )
  )

ggplot(df_0525_window,
       aes(x = day_of_water_year, y = snow_density,
           group = water_year, color = year_group)) +

  geom_line(alpha = 0.8) +

  # Piecewise empirical model
  geom_line(data = model_piecewise,
            aes(x = day_of_water_year, y = model_density, group = 1),
            inherit.aes = FALSE,
            linewidth = 1.2,
            color = "red") +

  scale_color_manual(
    values = c(
      "Other" = "grey75",
      "2011" = "#2259E0",
      "2012" = "#38E022"
    ),
    breaks = c("2011", "2012", "Other"),
    labels = c("2011", "2012", "Other years:\n 2005-2010,\n 2013-2025")
  ) +

  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +

  labs(
    title = "Bulk Snowpack Density by Water Year (2005–2025) at Deadman Hill SNOTEL",
    subtitle = "Two-part linear snowpack density equations (red line) derived from Mizukami and Perica (2008)",
    x = "Month (Water Year)",
    y = "Bulk Snow Density (kg/m³)",
    color = "Water Year"
  ) +
  theme_minimal()
```
**Figure 1.** Seasonal evolution of bulk snowpack density at the Deadman Hill SNOTEL site for water years 2005–2025, restricted to December 1 through April 30. Water years 2011 and 2012 are highlighted for comparison, while all remaining years are shown in gray. The red line shows a piecewise empirical density model using 𝜌=164.7+0.4" " 𝑡_𝑊𝑌from December 1 to March 1 and 𝜌=−90.7+2.1" " 𝑡_𝑊𝑌from March 1 to April 30 (1). 
      
# trend line created (NOT M-P)      
```{r}
# one fitted avg density line from Dec 1 - Mar 1, Mar 1 - Apr 30 ON TOP of the years 

# ----------------------------
# 1) Keep full water year data (2005–2025)
# ----------------------------
df_0525 <- df %>%
  filter(water_year >= 2005, water_year <= 2025) %>%
  filter(!is.na(snow_density)) %>%
  arrange(water_year, day_of_water_year)

# ----------------------------
# 2) Average density by day-of-water-year across years
# ----------------------------
avg_by_day <- df_0525 %>%
  group_by(day_of_water_year) %>%
  summarise(avg_density = mean(snow_density, na.rm = TRUE), .groups = "drop") %>%
  arrange(day_of_water_year)

# ----------------------------
# 3) Define key day-of-water-year cut points
# ----------------------------
dec1 <- 62
mar1 <- 152
apr30 <- 212

fit1_data <- avg_by_day %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= mar1)

fit2_data <- avg_by_day %>%
  filter(day_of_water_year >= mar1, day_of_water_year <= apr30)

fit1 <- lm(avg_density ~ day_of_water_year, data = fit1_data)
fit2 <- lm(avg_density ~ day_of_water_year, data = fit2_data)

# ----------------------------
# 4) One continuous piecewise prediction
# ----------------------------
pred_piecewise <- tibble(
  day_of_water_year = seq(dec1, apr30)
) %>%
  mutate(
    fit_density = if_else(
      day_of_water_year <= mar1,
      as.numeric(predict(fit1, newdata = .)),
      as.numeric(predict(fit2, newdata = .))
    )
  ) %>%
  filter(is.finite(fit_density))

# ----------------------------
# 5) Month breaks/labels
# ----------------------------
month_breaks <- c(62, 93, 124, 152, 183, 212)
month_labels <- c("Dec","Jan","Feb","Mar","Apr","Apr 30")

# ----------------------------
# 6) Plot (Dec 1–Apr 30 only)
# ----------------------------
df_0525_window <- df_0525 %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30) %>%
  mutate(
    year_group = case_when(
      water_year == 2011 ~ "2011",
      water_year == 2012 ~ "2012",
      TRUE ~ "Other"
    )
  )

ggplot(df_0525_window,
       aes(x = day_of_water_year, y = snow_density,
           group = water_year, color = year_group)) +

  geom_line(alpha = 0.8) +

  geom_line(data = pred_piecewise,
            aes(x = day_of_water_year, y = fit_density, group = 1),
            inherit.aes = FALSE,
            linewidth = 1,
            color = "red") +

  scale_color_manual(
    values = c(
      "Other" = "grey75",
      "2011" = "#2259E0",
      "2012" = "#38E022"
    ),
    breaks = c("2011", "2012", "Other"),
    labels = c("2011", "2012", "Other years:\n 2005-2010,\n 2013-2025")
  ) +

  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +

  labs(
    title = "Bulk Snowpack Density by Water Year (2005–2025) — Dec 1 to Apr 30",
    x = "Month (Water Year)",
    y = "Bulk Snow Density (kg/m³)",
    color = "Water Year"
  ) +
  theme_minimal()
```


# Density over the years, starting December 1
```{r, echo=FALSE}
# add a line for swe, density, and depth
library(dplyr)
library(lubridate)
library(ggplot2)

# ----------------------------
# Constants (non-leap reference)
# ----------------------------
dec1  <- 62
mar1  <- 152
apr30 <- 212

month_breaks <- c(62, 93, 124, 152, 183, 212)
month_labels <- c("Dec","Jan","Feb","Mar","Apr","Apr 30")

# ----------------------------
# 1) BASELINE: build 2005–2020 piecewise trend (Dec 1–Apr 30)
# ----------------------------
df_base <- df %>%
  filter(water_year >= 2005, water_year <= 2020) %>%
  filter(!is.na(snow_density)) %>%
  arrange(water_year, day_of_water_year)

avg_by_day_base <- df_base %>%
  group_by(day_of_water_year) %>%
  summarise(avg_density = mean(snow_density, na.rm = TRUE), .groups = "drop") %>%
  arrange(day_of_water_year)

fit1_data <- avg_by_day_base %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= mar1)

fit2_data <- avg_by_day_base %>%
  filter(day_of_water_year >= mar1, day_of_water_year <= apr30)

fit1 <- lm(avg_density ~ day_of_water_year, data = fit1_data)
fit2 <- lm(avg_density ~ day_of_water_year, data = fit2_data)

baseline_trend <- tibble(
  day_of_water_year = seq(dec1, apr30)
) %>%
  mutate(
    fit_density = if_else(
      day_of_water_year <= mar1,
      as.numeric(predict(fit1, newdata = .)),
      as.numeric(predict(fit2, newdata = .))
    )
  ) %>%
  filter(is.finite(fit_density))

# ----------------------------
# 2) TARGET YEARS: 2021–2025 lines (Dec 1–Apr 30 only)
# ----------------------------
df_2125 <- df %>%
  filter(water_year >= 2021, water_year <= 2025) %>%
  filter(!is.na(snow_density)) %>%
  arrange(water_year, day_of_water_year) %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30)

# ----------------------------
# 3) Plot 2021–2025 with 2005–2020 baseline trend overlaid
# ----------------------------
ggplot(df_2125,
       aes(x = day_of_water_year, y = snow_density,
           group = water_year, color = factor(water_year))) +

  geom_line(alpha = 0.85) +

  geom_line(data = baseline_trend,
            aes(x = day_of_water_year, y = fit_density, group = 1),
            inherit.aes = FALSE,
            linewidth = 1,
            color = "black") +

  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +

  labs(
    title = "Bulk Snowpack Density (2021–2025) with 2005–2020 Baseline Trend — Dec 1 to Apr 30",
    x = "Month (Water Year)",
    y = "Bulk Snow Density (kg/m³)",
    color = "Water Year"
  ) +
  theme_minimal()

# density stays in kg/m^3
```

# Change in Depth
1. compute the modeled density for WY 2024 using the two equations
2. use that modeled density plus observed SWE to calculate modeled snow depth
3. compute daily change in observed snow depth
4, compute daily change in modeled snow depth
5. plot the difference:
# Daily Change in Snow Depth
```{r}
# swe is in mm
# depth is in mm
library(dplyr)
library(ggplot2)

# ----------------------------
# Constants
# ----------------------------
dec1  <- 62
mar1  <- 152
apr30 <- 212
target_year <- 2024

month_breaks <- c(62, 93, 124, 152, 183)
month_labels <- c("Dec","Jan","Feb","Mar","Apr")

# ----------------------------
# 1) Extract WY 2024 data
# ----------------------------
df_year <- df %>%
  filter(water_year == target_year) %>%
  arrange(day_of_water_year) %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30)

# ----------------------------
# 2) Compute modeled density and modeled depth
# Dec 1–Mar 1:   density = 164.7 + 0.4 * t_WY
# Mar 1–Apr 30:  density = -90.7 + 2.1 * t_WY
# ----------------------------
df_model <- df_year %>%
  filter(!is.na(snow_density),
         !is.na(snow_depth),
         !is.na(snow_water_equivalent)) %>%
  mutate(
    modeled_density = case_when(
      day_of_water_year <= mar1 ~ 164.7 + 0.4 * day_of_water_year,
      day_of_water_year >  mar1 ~ -90.7 + 2.1 * day_of_water_year
    ),

    # observed depth already in cm
    snow_depth_cm = snow_depth,

    # SWE is in mm, modeled density is in kg/m3
    # resulting modeled depth is converted to cm
    modeled_depth_cm = (snow_water_equivalent / modeled_density) * 100,

    # density residual
    density_residual = snow_density - modeled_density
  ) %>%
  arrange(day_of_water_year) %>%
  mutate(
    # daily change in observed and modeled snow depth
    delta_obs_depth_cm   = snow_depth_cm - lag(snow_depth_cm),
    delta_model_depth_cm = modeled_depth_cm - lag(modeled_depth_cm),

    # observed minus modeled daily change
    delta_depth_residual_cm = delta_obs_depth_cm - delta_model_depth_cm
  )

# ----------------------------
# 3) Plot observed vs modeled density
# ----------------------------
ggplot(df_model, aes(x = day_of_water_year)) +
  geom_line(aes(y = snow_density), linewidth = 1) +
  geom_line(aes(y = modeled_density), linewidth = 1, linetype = "dashed", color = "red") +
  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +
  labs(
    title = "Bulk Snowpack Density vs Empirical Model (WY 2024)",
    x = "Month (Dec–Apr)",
    y = "Bulk Snow Density (kg/m³)"
  ) +
  theme_minimal()

# ----------------------------
# 4) Plot density residuals
# ----------------------------
ggplot(df_model, aes(x = day_of_water_year, y = density_residual)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_line(linewidth = 1) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +
  labs(
    title = "Density Residuals (Observed − Modeled) (WY 2024)",
    x = "Month (Dec–Apr)",
    y = "Residual Density (kg/m³)"
  ) +
  theme_minimal()

# ----------------------------
# 7) Plot daily change residual
# observed minus modeled
# ----------------------------
ggplot(df_model, aes(x = day_of_water_year, y = delta_depth_residual_cm)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_line(linewidth = 1) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +
  labs(
    title = "Daily Change in Snow Depth Residual (WY 2024)",
    x = "Month (Dec–Apr)",
    y = expression(Delta*" Snow Depth Residual (cm)")
  ) +
  theme_minimal()
```
**Figure 2.** Daily change in snow depth residual for water year 2024 at the Deadman Hill SNOTEL site, calculated as observed daily snow depth change minus modeled daily snow depth change. Modeled snow depth was derived from observed snow water equivalent and a piecewise empirical snow density model, with \(\rho = 164.7 + 0.4\,t_{WY}\) from December 1 to March 1 and \(\rho = -90.7 + 2.1\,t_{WY}\) from March 1 to April 30. Positive values indicate days when observed snow depth increased more than predicted, whereas negative values indicate days when snow depth decreased more than predicted.

# Density vs Depth Plots (2024)
```{r}
library(dplyr)
library(ggplot2)

# ----------------------------
# Constants
# ----------------------------
dec1  <- 62
mar1  <- 152
apr30 <- 212
target_year <- 2024

month_breaks <- c(62, 93, 124, 152, 183)
month_labels <- c("Dec","Jan","Feb","Mar","Apr")

# ----------------------------
# 1) Build training dataset (all years)
# ----------------------------
df_train <- df %>%
  filter(water_year >= 2005, water_year <= 2025) %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30) %>%
  filter(!is.na(snow_density),
         !is.na(snow_depth),
         !is.na(snow_water_equivalent)) %>%
  arrange(water_year, day_of_water_year) %>%
  group_by(water_year) %>%
  mutate(
    # observed daily change in snow depth (cm)
    delta_ds = snow_depth - lag(snow_depth)
  ) %>%
  ungroup() %>%
  mutate(
    season_period = case_when(
      day_of_water_year <= mar1 ~ "Dec1-Mar1",
      day_of_water_year >  mar1 ~ "Mar1-Apr30"
    ),

    # seasonal mean density from M-P equations
    rho_mean_mp = case_when(
      day_of_water_year <= mar1 ~ 164.7 + 0.4 * day_of_water_year,
      day_of_water_year >  mar1 ~ -90.7 + 2.1 * day_of_water_year
    ),

    # density departure from mean
    delta_rho = snow_density - rho_mean_mp
  ) %>%
  filter(!is.na(delta_ds), !is.na(delta_rho))

# ----------------------------
# 2) Fit correction function f(Δds)
# separate linear model for each season
# ----------------------------
fit_early <- lm(delta_rho ~ delta_ds,
                data = df_train %>% filter(season_period == "Dec1-Mar1"))

fit_late  <- lm(delta_rho ~ delta_ds,
                data = df_train %>% filter(season_period == "Mar1-Apr30"))

summary(fit_early)
summary(fit_late)

# ----------------------------
# 3) Apply model to WY 2024
# ----------------------------
df_2024 <- df %>%
  filter(water_year == target_year) %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30) %>%
  filter(!is.na(snow_density),
         !is.na(snow_depth),
         !is.na(snow_water_equivalent)) %>%
  arrange(day_of_water_year) %>%
  mutate(
    delta_ds = snow_depth - lag(snow_depth),

    season_period = case_when(
      day_of_water_year <= mar1 ~ "Dec1-Mar1",
      day_of_water_year >  mar1 ~ "Mar1-Apr30"
    ),

    rho_mean_mp = case_when(
      day_of_water_year <= mar1 ~ 164.7 + 0.4 * day_of_water_year,
      day_of_water_year >  mar1 ~ -90.7 + 2.1 * day_of_water_year
    ),

    delta_rho_obs = snow_density - rho_mean_mp
  )

# predict correction from Δds
df_2024 <- df_2024 %>%
  mutate(
    delta_rho_pred = case_when(
      season_period == "Dec1-Mar1"  ~ predict(fit_early, newdata = .),
      season_period == "Mar1-Apr30" ~ predict(fit_late,  newdata = .)
    ),

    rho_model = rho_mean_mp + delta_rho_pred
  )

# ----------------------------
# 4) Plot observed density vs seasonal mean vs corrected model
# ----------------------------
ggplot(df_2024, aes(x = day_of_water_year)) +
  geom_line(aes(y = snow_density), linewidth = 1) +
  geom_line(aes(y = rho_mean_mp), linewidth = 1, linetype = "dashed", color = "red") +
  geom_line(aes(y = rho_model), linewidth = 1, color = "blue") +
  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +
  labs(
    title = "Observed and Modeled Bulk Snow Density (WY 2024)",
    x = "Month (Dec–Apr)",
    y = "Bulk Snow Density (kg/m³)"
  ) +
  theme_minimal()

# Scatterplot Version wy24
library(dplyr)
library(ggplot2)

# ----------------------------
# Constants
# ----------------------------
dec1  <- 62
apr30 <- 212
target_year <- 2024

# ----------------------------
# Build WY 2024 dataset
# ----------------------------
df_scatter <- df %>%
  filter(water_year == target_year) %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30) %>%
  filter(!is.na(snow_depth), !is.na(snow_density)) %>%
  arrange(day_of_water_year) %>%
  mutate(
    delta_depth   = snow_depth - lag(snow_depth),         # cm
    delta_density = snow_density - lag(snow_density)      # kg/m^3
  ) %>%
  filter(!is.na(delta_depth), !is.na(delta_density))

# ----------------------------
# Scatterplot: change in density vs change in depth
# ----------------------------
ggplot(df_scatter, aes(x = delta_depth, y = delta_density)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey50") +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 1) +
  labs(
    title = "Daily Change in Snow Density vs Daily Change in Snow Depth (WY 2024)",
    x = expression(Delta*" Snow Depth (cm)"),
    y = expression(Delta*" Bulk Snow Density (kg/"*m^3*")")
  ) +
  theme_minimal()
```

# Map
```{r}
library(ggplot2)
library(maps)
library(patchwork)
library(grid)

# ----------------------------
# Coordinates
# ----------------------------
site_lat <- 40 + 48/60
site_lon <- -(105 + 46/60)

site_df <- data.frame(
  lon = site_lon,
  lat = site_lat
)

us_map <- map_data("state")

co_box <- data.frame(
  xmin = -109.05,
  xmax = -102.05,
  ymin = 37.00,
  ymax = 41.00
)

# ----------------------------
# Scale bar setup
# ----------------------------
km_per_deg_lon <- 111.32 * cos(site_lat * pi / 180)

bar_km <- 50
bar_deg <- bar_km / km_per_deg_lon

bar_x0 <- -108.6
bar_x1 <- bar_x0 + bar_deg
bar_y  <- 37.25

# ----------------------------
# Locator map (U.S.)
# ----------------------------
p_us <- ggplot() +
  geom_polygon(
    data = us_map,
    aes(x = long, y = lat, group = group),
    fill = "grey92",
    color = "grey45",
    linewidth = 0.2
  ) +
  geom_rect(
    data = co_box,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = NA,
    color = "red",
    linewidth = 0.9,
    linetype = "solid"
  ) +
  # ✅ ADD POINT TO LOCATOR MAP
  geom_point(
    data = site_df,
    aes(x = lon, y = lat),
    size = 2.5
  ) +
  coord_fixed(xlim = c(-125, -66), ylim = c(24, 50)) +
  labs(
    title = "Locator map",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

# ----------------------------
# Colorado map
# ----------------------------
p_co <- ggplot() +
  geom_rect(
    data = co_box,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "grey98",
    color = "black",
    linewidth = 0.8
  ) +
  geom_point(
    data = site_df,
    aes(x = lon, y = lat),
    size = 3,
    color = "red"
  ) +
  # ✅ SHIFTED LABEL DOWN
  annotate(
    "text",
    x = site_lon + 0.25,
    y = site_lat - 0.20,
    label = "Deadman Hill SNOTEL\n40°48′N, 105°46′W\nElevation: 3,121 m",
    hjust = 0,
    size = 3.2
  ) +
  # North arrow
  annotate(
    "segment",
    x = -102.6, xend = -102.6,
    y = 37.4,  yend = 37.9,
    linewidth = 0.6,
    arrow = arrow(length = unit(0.15, "in"))
  ) +
  annotate(
    "text",
    x = -102.6, y = 38.02,
    label = "N",
    size = 4
  ) +
  # Scale bar
  annotate(
    "segment",
    x = bar_x0, xend = bar_x1,
    y = bar_y, yend = bar_y,
    linewidth = 1
  ) +
  annotate(
    "segment",
    x = bar_x0, xend = bar_x0,
    y = bar_y - 0.03, yend = bar_y + 0.03,
    linewidth = 1
  ) +
  annotate(
    "segment",
    x = bar_x1, xend = bar_x1,
    y = bar_y - 0.03, yend = bar_y + 0.03,
    linewidth = 1
  ) +
  annotate(
    "text",
    x = (bar_x0 + bar_x1) / 2,
    y = bar_y + 0.08,
    label = "50 km",
    size = 3
  ) +
  coord_fixed(xlim = c(-109.05, -102.05), ylim = c(37.00, 41.00)) +
  labs(
    title = "Deadman Hill SNOTEL Site",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

# ----------------------------
# Combine
# ----------------------------
final_map <- p_us + p_co +
  plot_annotation(title = "Deadman Hill SNOTEL Site Map")

final_map
```

# 2011 and 2012 plots with MP08
```{r}
# Line plot
library(dplyr)
library(ggplot2)

# ----------------------------
# Constants
# ----------------------------
dec1  <- 62
mar1  <- 152
apr30 <- 212

month_breaks <- c(62, 93, 124, 152, 183)
month_labels <- c("Dec","Jan","Feb","Mar","Apr")

# ----------------------------
# 1) Extract WY 2011 and WY 2012
# ----------------------------
df_lines <- df %>%
  filter(water_year %in% c(2011, 2012)) %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30) %>%
  filter(!is.na(snow_density)) %>%
  arrange(water_year, day_of_water_year) %>%
  mutate(
    wy_label = case_when(
      water_year == 2011 ~ "2011",
      water_year == 2012 ~ "2012"
    )
  )

# ----------------------------
# 2) Piecewise empirical trend line
# ----------------------------
trend_line <- tibble(
  day_of_water_year = seq(dec1, apr30)
) %>%
  mutate(
    modeled_density = case_when(
      day_of_water_year <= mar1 ~ 164.7 + 0.4 * day_of_water_year,
      day_of_water_year >  mar1 ~ -90.7 + 2.1 * day_of_water_year
    )
  )

# ----------------------------
# 3) Plot
# ----------------------------
# ----------------------------
# Constants
# ----------------------------
dec1  <- 62
apr30 <- 212

# ----------------------------
# Build scatterplot data for WY 2011 and WY 2012
# ----------------------------
df_scatter <- df %>%
  filter(water_year %in% c(2011, 2012)) %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30) %>%
  filter(!is.na(snow_depth), !is.na(snow_density)) %>%
  arrange(water_year, day_of_water_year) %>%
  group_by(water_year) %>%
  mutate(
    delta_depth   = snow_depth - lag(snow_depth),      # cm
    delta_density = snow_density - lag(snow_density),  # kg/m^3
    wy_label = as.factor(water_year)
  ) %>%
  ungroup() %>%
  filter(!is.na(delta_depth), !is.na(delta_density))

ggplot() +
  geom_line(
    data = df_lines,
    aes(x = day_of_water_year, y = snow_density, color = wy_label),
    linewidth = 1.1
  ) +
  geom_line(
    data = trend_line,
    aes(x = day_of_water_year, y = modeled_density),
    color = "red",
    linetype = "dashed",
    linewidth = 1.1
  ) +
  scale_color_manual(
    values = c("2011" = "blue", "2012" = "black")
  ) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +
  labs(
    title = "Observed Bulk Snow Density for WY 2011 and WY 2012",
    x = "Month (Dec–Apr)",
    y = expression("Bulk Snow Density (kg/"*m^3*")"),
    color = "Water Year"
  ) +
  theme_minimal()

ggplot(df_scatter, aes(x = delta_depth, y = delta_density, color = wy_label)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey50") +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_manual(
    values = c("2011" = "blue", "2012" = "black")
  ) +
  facet_wrap(~wy_label) +
  labs(
    title = "Daily Change in Snow Density vs Daily Change in Snow Depth",
    x = expression(Delta*" Snow Depth (cm)"),
    y = expression(Delta*" Bulk Snow Density (kg/"*m^3*")")
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

# 2011 and 2012 plots with own trend line
```{r}
library(dplyr)
library(ggplot2)
library(tibble)

# ----------------------------
# 1) Keep full water year data (2005–2025)
# ----------------------------
df_0525 <- df %>%
  filter(water_year >= 2005, water_year <= 2025) %>%
  filter(!is.na(snow_density)) %>%
  arrange(water_year, day_of_water_year)

# ----------------------------
# 2) Average density by day-of-water-year across years
# ----------------------------
avg_by_day <- df_0525 %>%
  group_by(day_of_water_year) %>%
  summarise(avg_density = mean(snow_density, na.rm = TRUE), .groups = "drop") %>%
  arrange(day_of_water_year)

# ----------------------------
# 3) Define key day-of-water-year cut points
# ----------------------------
dec1 <- 62
mar1 <- 152
apr30 <- 212

fit1_data <- avg_by_day %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= mar1)

fit2_data <- avg_by_day %>%
  filter(day_of_water_year >= mar1, day_of_water_year <= apr30)

fit1 <- lm(avg_density ~ day_of_water_year, data = fit1_data)
fit2 <- lm(avg_density ~ day_of_water_year, data = fit2_data)

# ----------------------------
# 4) One continuous piecewise prediction
# ----------------------------
pred_piecewise <- tibble(
  day_of_water_year = seq(dec1, apr30)
) %>%
  mutate(
    fit_density = if_else(
      day_of_water_year <= mar1,
      as.numeric(predict(fit1, newdata = .)),
      as.numeric(predict(fit2, newdata = .))
    )
  ) %>%
  filter(is.finite(fit_density))

# ----------------------------
# 5) Month breaks/labels
# ----------------------------
month_breaks <- c(62, 93, 124, 152, 183)
month_labels <- c("Dec","Jan","Feb","Mar","Apr")

# ----------------------------
# 6) Extract only WY 2011 and WY 2012 for line plot
# ----------------------------
df_lines <- df_0525 %>%
  filter(water_year %in% c(2011, 2012)) %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30) %>%
  mutate(
    wy_label = case_when(
      water_year == 2011 ~ "2011",
      water_year == 2012 ~ "2012"
    )
  )

# ----------------------------
# 7) Plot
# ----------------------------
ggplot() +
  geom_line(
    data = df_lines,
    aes(x = day_of_water_year, y = snow_density, color = wy_label),
    linewidth = 1.2
  ) +
  geom_line(
    data = pred_piecewise,
    aes(x = day_of_water_year, y = fit_density),
    color = "red",
    linewidth = 1.1
  ) +
  scale_color_manual(
    values = c("2011" = "blue", "2012" = "black")
  ) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +
  labs(
    title = "Observed Bulk Snow Density for WY 2011 and WY 2012 WITHOUT MP08",
    x = "Month (Dec–Apr)",
    y = expression("Bulk Snow Density (kg/"*m^3*")"),
    color = "Water Year"
  ) +
  theme_minimal()

library(dplyr)
library(ggplot2)
library(tibble)

# ----------------------------
# Reuse fitted trend from all years
# ----------------------------
df_fit_lookup <- pred_piecewise %>%
  select(day_of_water_year, fit_density)

library(dplyr)
library(ggplot2)

dec1 <- 62
apr30 <- 212

df_scatter <- df %>%
  filter(water_year %in% c(2011, 2012)) %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30) %>%
  filter(!is.na(snow_depth), !is.na(snow_density)) %>%
  arrange(water_year, day_of_water_year) %>%
  group_by(water_year) %>%
  mutate(
    delta_depth   = snow_depth - lag(snow_depth),
    delta_density = snow_density - lag(snow_density),
    wy_label = as.factor(water_year)
  ) %>%
  ungroup() %>%
  filter(!is.na(delta_depth), !is.na(delta_density))

ggplot(df_scatter, aes(x = delta_depth, y = delta_density, color = wy_label)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey50") +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_manual(
    values = c("2011" = "blue", "2012" = "black")
  ) +
  labs(
    title = "Daily Change in Snow Density vs Daily Change in Snow Depth",
    x = expression(Delta*" Snow Depth (cm)"),
    y = expression(Delta*" Bulk Snow Density (kg/"*m^3*")"),
    color = "Water Year"
  ) +
  theme_minimal()
```

# Depth plots for 2011 and 2012
```{r}

```

# Comparison
  # should not go over 1500 mm of SWE. Using MP08 and average density to look at depth
  # use depth alone to get an accurate density 
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(tibble)

# ----------------------------
# Constants
# ----------------------------
dec1  <- 62
mar1  <- 152
apr30 <- 212

month_breaks <- c(62, 93, 124, 152, 183)
month_labels <- c("Dec","Jan","Feb","Mar","Apr")

# ----------------------------
# 1) Build fitted average-density trend from 2005–2025
# ----------------------------
df_0525 <- df %>%
  filter(water_year >= 2005, water_year <= 2025) %>%
  filter(!is.na(snow_density)) %>%
  arrange(water_year, day_of_water_year)

avg_by_day <- df_0525 %>%
  group_by(day_of_water_year) %>%
  summarise(avg_density = mean(snow_density, na.rm = TRUE), .groups = "drop") %>%
  arrange(day_of_water_year)

fit1_data <- avg_by_day %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= mar1)

fit2_data <- avg_by_day %>%
  filter(day_of_water_year >= mar1, day_of_water_year <= apr30)

fit1 <- lm(avg_density ~ day_of_water_year, data = fit1_data)
fit2 <- lm(avg_density ~ day_of_water_year, data = fit2_data)

pred_piecewise <- tibble(
  day_of_water_year = seq(dec1, apr30)
) %>%
  mutate(
    rho_avgfit = if_else(
      day_of_water_year <= mar1,
      as.numeric(predict(fit1, newdata = .)),
      as.numeric(predict(fit2, newdata = .))
    )
  )

# ----------------------------
# 2) Join modeled density to WY 2011 and WY 2012
# ----------------------------
df_compare <- df %>%
  filter(water_year %in% c(2011, 2012)) %>%
  filter(day_of_water_year >= dec1, day_of_water_year <= apr30) %>%
  filter(!is.na(snow_depth), !is.na(snow_water_equivalent)) %>%
  left_join(pred_piecewise, by = "day_of_water_year") %>%
  mutate(
    # snow_depth is in cm → convert to meters
    depth_m = snow_depth / 100,

    # modeled SWE in mm
    swe_model_avgfit = rho_avgfit * depth_m,

    # residuals
    swe_resid_avgfit = swe_model_avgfit - snow_water_equivalent,

    # absolute error
    abs_err_avgfit = abs(swe_resid_avgfit),

    wy_label = as.factor(water_year)
  )

# ----------------------------
# 3) Summary metrics by year
# ----------------------------
swe_summary <- df_compare %>%
  group_by(water_year) %>%
  summarise(
    mean_bias = mean(swe_resid_avgfit, na.rm = TRUE),
    mae = mean(abs_err_avgfit, na.rm = TRUE),
    rmse = sqrt(mean(swe_resid_avgfit^2, na.rm = TRUE)),
    .groups = "drop"
  )

print(swe_summary)

# ----------------------------
# 4) Time series plot for 2011 and 2012
# ----------------------------
df_long <- df_compare %>%
  select(
    water_year, day_of_water_year,
    snow_water_equivalent, swe_model_avgfit
  ) %>%
  rename(
    `Observed SWE` = snow_water_equivalent,
    `Modeled SWE`  = swe_model_avgfit
  ) %>%
  pivot_longer(
    cols = c(`Observed SWE`, `Modeled SWE`),
    names_to = "series",
    values_to = "SWE_mm"
  )

ggplot(df_long,
       aes(x = day_of_water_year, y = SWE_mm, color = series)) +
  geom_line(linewidth = 1) +
  facet_wrap(~water_year, ncol = 1) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +
  labs(
    title = "Observed vs Modeled SWE for WY 2011 and WY 2012",
    x = "Month (Dec–Apr)",
    y = "SWE (mm)",
    color = NULL
  ) +
  theme_minimal()

# ----------------------------
# 5) Scatterplot: modeled vs observed SWE
# ----------------------------
ggplot(df_compare,
       aes(x = snow_water_equivalent, y = swe_model_avgfit, color = wy_label)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  scale_color_manual(
    values = c("2011" = "blue", "2012" = "black")
  ) +
  labs(
    title = "Modeled SWE from Observed Depth and Fitted Average Density",
    x = "Observed SWE (mm)",
    y = "Modeled SWE (mm)",
    color = "Water Year"
  ) +
  theme_minimal()

# ----------------------------
# 6) Optional: residuals through time
# ----------------------------
ggplot(df_compare,
       aes(x = day_of_water_year, y = swe_resid_avgfit, color = wy_label)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(linewidth = 1) +
  facet_wrap(~water_year, ncol = 1) +
  scale_color_manual(
    values = c("2011" = "blue", "2012" = "black")
  ) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels,
    limits = c(dec1, apr30),
    expand = c(0, 0)
  ) +
  labs(
    title = "Modeled SWE Residuals for WY 2011 and WY 2012",
    x = "Month (Dec–Apr)",
    y = "Modeled SWE - Observed SWE (mm)",
    color = "Water Year"
  ) +
  theme_minimal()
```

